---
title: "Exercise 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. [0.5 points]
We load the package $\texttt{jpeg}$ (install first if not already done) and take a look at the function documentation of $\texttt{readJPEG}$. We see that we have to specify the path where the picture is located and that an array of dimension height x width x 3 is returned. 

```{r, warning=FALSE}
library(jpeg)
img <- readJPEG( "C:/Users/leaz9/OneDrive/Dokumente/StatLearn WS22/ColorfulBird.jpg")
dim(img)
```

## 2. [1.5 points]
When we want to plot the picture, we simply plot a coloured point for every pixel. Therefore, having the correct $x$ and $y$ positions of the pixels on the axis is important. 

The `img`-array has stored the pixel values starting in the top left corner of the picture and then moving columnwise to the bottom right corner. 
Thus, the $526 \times 800$ matrix `img[ , , 1]` contains the values of the red color channel from top left to bottom right,
and similar for green and blue values. 
We then get the vector of red values by running `as.vector(img[, , 1])`. Here, the matrix is flattened columnwise, so the first 526 components are the $y$-values for $x = 1$ in the order $y =256$ to $y = 1$,  the next 526 components are the $y$-values for $x = 2$ and so on. 
We can then build the dataframe as follows: 

```{r}

imgRGB <- data.frame(
  x = rep(1:800, each = 526),
  y = rep(526:1, 800),
  R = as.vector(img[,,1]),
  G = as.vector(img[,,2]),
  B = as.vector(img[,,3])
)

dim(imgRGB)
```

Now we can plot the image by running

```{r, fig.dim = c(12,10)}
plot( imgRGB$x, imgRGB$y, col = rgb(imgRGB[c("R", "G", "B")]))

```


The function `rgb()` converts the three primary colours into the corresponding colour.

## 3. [0.5 points]

We want to cluster the pixel colors into 10 clusters with `kmeans()`. Here, each pixel is understood as an observation of three variables.

```{r}
set.seed(2021)
img_km <- kmeans(imgRGB[, c("R", "G", "B")], centers = 10, nstart = 10, iter.max = 100)

```
We now have the folowing 10 cluster centers

```{r}
img_km$centers
```
We can have a look at the colours that are represented by the cluster centers as follows.
```{r, fig.dim = c(6,4)}
scales::show_col(rgb(img_km$centers))
```

## 4. [1 point]

The cluster belongings of the pixels can be accessed with `img_km$cluster`. It's a vector with entries in $\{ 1, 2, \ldots, 10\}$ representing the cluster number that the respective pixel belongs to. Since we have $526 \cdot 800 = 420800$ pixels, the vector with cluster belongings is also of that length.
We want each pixel to be coloured in the colour of the cluster center that it belongs to. These new pixel colors are
```{r}
km_colours <- rgb(img_km$centers[img_km$cluster, ])
```
Now we plot

```{r, fig.dim = c(12,10)}
plot( imgRGB$x, imgRGB$y, col = km_colours)
```

## 5. [0.5 points]

For the compressed file, we need to store a matrix (or vector, you don't have to transform it to a matrix when you can remember the correct order of pixels within the vector) with the cluster belongings of the pixels and the cluster centers, i.e.

```{r}

pixel2clust <- matrix(img_km$cluster, ncol = 526)

km_centers <- img_km$centers

compr_size <- object.size(km_centers) + object.size(pixel2clust)
compr_size

```
The original image and the ratio of

```{r}
orig_size <- object.size(img)
orig_size

as.numeric(orig_size/compr_size)
```

So the memory that the original image uses is 6 times the memory that the compressed file uses.



