---
title: " "
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 2

First of all, we import the data and specify what the columns represent. 
We further take a look at the dimension of the data.
```{r}
frenchfood <- read.csv("~/StatLearn WS22/Übungszettel_WS2122/data/french-food.txt", sep="", 
                       row.names = 1)
head(frenchfood)
colnames(frenchfood) <- c("bread", "vegetables", "fruit", "meat", "poultry", "milk", "wine")
dim(frenchfood)

```




### a) [ 1 point]
From the formula in the lecture, we know that for $d = 7$ we have $\Delta \geq 0$ when $p \leq 3$, 
 so an orthogonal factor model with 3 or less factors makes sense.

**Eigenvalue method:**
```{r}
cor_food <- cor(frenchfood)
eig_food <- eigen(cor_food)

eig_food$values
sum(eig_food$values > 1)
```

The first two eigenvalues of the correlation matrix are greater than 1. 
The criterion thus suggests to use 2 factors.

**Screeplot method:** Here, the screeplot just plots the index of the eigenvalue vs the eigenvalue (of the correlation matrix).

```{r}
plot(1:7, eig_food$values, type = "o", xlab = "j", ylab = expression(lambda[j]))
```

This does not give a clear answer, either  2 or 3 factors might be considered.

**Formal test based on normality assumption:** we can compute p-values for models with 2 or 3 factors.
```{r}
sapply(2:3, 
       function(nf){ 
         paste( "pvalue for ", nf , "factors:" , factanal(frenchfood, factors = nf )$PVAL)})

```

The formal test still rejects the null hypothesis $( p < 0.02)$. But 3 is the maximal 
amount of factors that we may use for $d = 7$ variables.

### b) [0.5 points]

```{r}
fa_ml <- factanal(frenchfood, factors = 3, rotation = "varimax", scores = "regression")
fa_ml

```


### c) [1.5 points]
```{r, fig.dim=c(8, 8)}
par(mfrow=c(2,2))
plot(fa_ml$loadings[,1:2], xlim=c(-1,1),ylim=c(-1,1))
text(fa_ml$loadings[,1:2], label=colnames(frenchfood), pos=2)
abline(v=0, lty=2)
abline(h=0, lty=2)

plot(fa_ml$loadings[,c(1,3)], xlim=c(-1,1),ylim=c(-1,1))
text(fa_ml$loadings[,c(1,3)], label=colnames(frenchfood), pos=2)
abline(v=0, lty=2)
abline(h=0, lty=2)

plot(fa_ml$loadings[,2:3], xlim=c(-1,1),ylim=c(-1,1))
text(fa_ml$loadings[,2:3], label=colnames(frenchfood), pos=2)
abline(v=0, lty=2)
abline(h=0, lty=2)

fa_ml$loadings

```


Vegetables, fruit, meat and poultry are loaded highest on factor 1.

Bread and milk are loaded highest on factor 2, vegetables are also loaded quite
high on factor 2.

Wine is loaded highest on factor 3.

**Interpretation:**
Factor 1 corresponds to food that is relatively pricey compared 
 to the amount of energy it provides.
Factor 2 corresponds to cheap foods that are filling.
Factor 3 only corresponds to wine, a luxury food, not providing any energy.

 

### d) [1 point]
```{r}
par(mfrow=c(1,1))
plot(fa_ml$scores[,1:2], xlim=c(-2,2),ylim=c(-2,2))
text(fa_ml$scores[,1:2], label=rownames(frenchfood), pos=2)
abline(v=0, lty=2)
abline(h=0, lty=2)
```


The families with 5 kids have high factor scores for factor 2,
i.e. they buy more bread and milk than families with less children do.
Probably because it is an easy and affordable way to feed a lot of children.
Further: all manager families have positive scores on factor 1. (more money spend on the
pricier foods)


