---
title: " "
author:   " "
date: " "
output:  
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR)
```

## a) [0.5 points]
First, we load the library and have a look at the documentation of the dataset.
```{r, echo = FALSE}
library(ISLR)
?NCI60
```

 The format is a list containing two elements: data and labs.
 'data' contains the expression levels on 6839 genes from 64 cancer cell lines; 
 'labs' is the corresponding cancer type (of the cancer cell lines)

```{r}
nci_labs <- NCI60$labs
nci_data <- NCI60$data
dim(nci_data)
```

Running $\texttt{head()}$ is not informative because it has more than 6000 columns. 
To be precise, we have 6830 gene expression measurements of 64 cancer cell lines.
Instead, we have a look at the first 10 observations of the first five observed variables.
```{r}
nci_data[1:10, 1:5]
```


## b) [0.5 points]
With the function $\texttt{table()}$ we can count how often which cancer type appears in the 
vector with the cancer types. Then we extract the names of those that appear more often than 3 times. 

```{r}
table(nci_labs)

which(table(nci_labs) > 3)        
names(which(table(nci_labs)>3))
chosen_rows <- nci_labs %in% names(which(table(nci_labs)>3))   
chosen_rows  
```
The last vector specifies for each row whether it is kept or not.
```{r}
nci_red <- nci_data[chosen_rows,]
dim(nci_red)
```


## c) [0.5 points]

Since we cannot have a look at all means and variances seperately ( too many dimensions), 
we just have a look at their summary statistics.
```{r}
summary(apply(nci_red, 2, var))
summary(apply(nci_red, 2, mean))
```


The variances range between 0.03 and 12. That's a factor of 400, so 
we should better scale the variables to have unit variance when performing PCA.

## d)  [0.5 points]
As seen in c), we should set $\texttt{scale = TRUE}$ in $\texttt{prcomp()}$
to scale the observations (centering is done per default).

```{r}
pc_out <- prcomp(nci_red, scale. = TRUE)

smry_pc <- summary(pc_out)
smry_pc
```

## e) [0.5 points each]
The importance matrix is one of the values in the list that is returned when running 
$\texttt{summary()}$ on the $\texttt{prcomp}$-object. One can access it with $ \texttt{smry_pc\$importance}$.
The cumulative proportion is the third row within that matrix. 
```{r}
smry_pc$importance[3, ]
which( smry_pc$importance[3, ] >= .8)[1]

```
For example, the first criterion would suggest to use 31 PCs if we would want to explain 80 \% of total variance.


For the second criterion, we need the average proportion of explained variance, which is the mean of
the second row of the importance matrix: 

```{r}
# average prop of explained variance
mean(smry_pc$importance[2, ])
# those with larger than average proportion of explained variance
which(smry_pc$importance[2,] >= mean(smry_pc$importance[2, ]))

```

The second criterion would choose 17 PCs.

The third criterion is the screeplot, which can be generated with

```{r}
screeplot(pc_out, type = "l", npcs = 30)
```

It would suggest to use 3 PCs. 

## f) [0.5 points]
We first define a function that returns a color for each element of a vector: 

```{r}
Cols <- function(vec){
  cols <- rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))]) 
}
```

The scores of the first two PCs are returned in the list element $x$ of the $\texttt{prcomp}$ object. 
```{r, fig.dim = c(6,4)}
plot(pc_out$x[,1:2], col = Cols(nci_labs[chosen_rows]), pch = 19,
     xlab="Z1",ylab="Z2")
legend("bottomright",  legend = unique(nci_labs[chosen_rows]), col = Cols(unique(nci_labs[chosen_rows])), 
       pch = 19, cex = 0.6)

```

For $\texttt{ggplot()}$ we define a dataframe with the scores and the corresponding cancer type first.

```{r, fig.dim=c(6,4)}
library(ggplot2)
df <- data.frame(pc_out$x[, 1:2])
df$CancerType <- nci_labs[chosen_rows]
ggplot(df, aes( x = PC1, y  = PC2, color = CancerType)) +
  geom_point(size = 2) +
   theme_bw()   # for black and white color theme
```




